---
title: "PUBH 7462 HW2"
author: "Anthony Johnson"
date: "2/10/2022"
output: github_document
---

```{r, message=FALSE}
library(tidyverse)
#Working directory for .RMD
knitr::opts_knit$set(echo = TRUE,
root.dir = rprojroot::find_rstudio_root_file())
#Controlling figure output in markdown
knitr::opts_chunk$set(
# fig.height = 4
fig.width = 6,
# fig.asp = .5,
out.width = "90%",
# out.height =
fig.align = "center",
cache = FALSE,
echo = TRUE
)
#Set Theme for ggplot2
theme_set(theme(plot.title = element_text(hjust = 0.5), legend.position = "bottom"))
#Set Scientific notation output for knitr
options(scipen = 999)
```

# 3.1 Data Exploration & Cleaning (10pts)

```{r, echo=FALSE, eval=FALSE}
library(DataExplorer)
library(readr)
plot_bar(brfss_smart_2010)
```


```{r, message=F}
brfss_smart_2010 <- read_csv("data/brfss_smart_2010.csv")
tidy_brfss <- brfss_smart_2010 %>%
  janitor::clean_names() %>% # Giving all variables clean, uniform names
  filter(topic %in% "Overall Health") %>%  # Only want Overall Health topic
  mutate(locationdesc = str_remove_all(string =  locationdesc, pattern =  "County")) %>% # Remove the county after each county name because it is redundant
  separate(locationdesc, c("state", "county"), sep = " - ") %>% # Separating state and county into their own columns
  rename(prop_responses = data_value) %>%
  select(year, state, county, response, sample_size, prop_responses) %>%
  mutate(state = factor(state), 
         response = factor(response, levels = c("Poor", "Fair", "Good", "Very good", "Excellent") ,
                           ordered = TRUE)) # Making variables into factors to make them easier to work with
```

# 3.2 Data Description (10pts)

There are `r nrow(tidy_brfss)` observations and `r ncol(tidy_brfss)` variables. An observation is a overall health response from a specific location. The variable `year` describes the year the data was collected (2002-2010). The variable `state` describes which state and `county` describes the county where the question was asked. The variable `response` is how the participants responded to the question of "How is your general health?", and is a factor variable with levels: "Excellent", "Very Good", "Good", "Fair" and "Poor." The variable `sample_size` is the number of people who gave each of those responses at each location. The variable `prop_responses` is the proportion of each response at each location.

# 3.3 Do Data Science (50pts)

## 3.3.1 In the year 2004, which states were observed at 6 locations? (10pts)
```{r}
tidy_brfss %>%
  filter(year %in% 2004) %>%
  distinct(state, county) %>%
  count(state) %>%
  filter(n %in% 6)
```

The states observed at 6 locations in 2004 were Colorado, Connecticut, Maryland, New Mexico, South Carolina, Texas, Utah, and Vermont.

## 3.3.2 Make a “spaghetti plot” that shows the number of observed locations in each state from 2002 to 2010. Which state has the highest mean number of locations over this period? (10pts)

```{r, warning=F, message=F}
tidy_brfss %>%
  distinct(year, state, county) %>%
  group_by(year, state) %>%
  summarize(n = n()) %>% 
  ggplot(aes(x = year, y = n, color = fct_reorder2(state, year, n, .desc = T))) +
  geom_smooth(se = F) + 
  labs(title = "Number of Observed Locations in Each State from 2002 to 2010", x = "Year", y = "Number of Locations", color = "State") +
  theme(legend.key.size = unit(0.1, "in"))
```
```{r, warning=F, message=F}
(mean_locations <- tidy_brfss %>%
  distinct(year, state, county) %>%
  group_by(year, state) %>%
  summarize(n = n()) %>%
  group_by(state) %>%
  summarise(Avg = mean(n)) %>%
  arrange(-Avg))
```




From the plot it appears that Florida and New Jersey are the top two states in terms of locations both at the end and over time (we used geom_smooth, so we plotted the conditional mean). Florida starts out low, but shoots up over time. New Jersey, North Carolina, and Washington are pretty steadily at the top in terms of numbers of locations over time. Most of the rest of the states are bunched together and the final order is labeled in the legend. In the above table, we see that New Jersey has the most average locations with `r round(mean_locations$Avg[1], digits = 2)`. Florida has the second most with `r round(mean_locations$Avg[2], digits = 2)`. 




## 3.3.3 Make a table showing, for the years 2002, 2006, and 2010, the mean and standard deviation of sample size and proportion of Excellent, Good, and Poor responses across locations in MN. (15pts)

```{r, message=F}
library(gt)
tidy_brfss %>%
  filter(year %in% c(2002, 2006, 2010), state %in% "MN") %>%
  group_by(year, response) %>%
  summarize(across(c(sample_size, prop_responses), list(mean = mean,sd = sd))) %>% # Find mean and sd of each value 
  gt() %>%
  tab_header("Sample Size and Proportion of Responses Across Locations in MN") %>%
  cols_label(response = "Response", sample_size_mean = "Mean of Sample Size", sample_size_sd = "Standard Deviation of Sample Size", prop_responses_mean = "Mean of Proportion", prop_responses_sd = "Standard Deviation of Proportion" ) %>%
  cols_align("center") %>%
  fmt_number(columns = c(sample_size_mean, sample_size_sd, prop_responses_mean, prop_responses_sd), decimals = 2) %>%
  as_raw_html() # Trying to fix weird output in github document
```

This table shows the mean and standard deviation for the sample size and proportion of each response. It is grouped by the years 2002, 2006, and 2010.


## 3.3.4 Create a ggplot that communicates the results/trends from the table above and stands on its own (15pts)




```{r, warning=F, message=F}
tidy_brfss %>%
  filter(year %in% c(2002, 2006, 2010), state %in% "MN") %>%
  group_by(year, response) %>%
  summarize(across(c(sample_size, prop_responses), list(mean = mean,sd = sd))) %>%
  ggplot(aes(x = prop_responses_mean, y = sample_size_mean, color = response)) +
  geom_point(size = 1) +
  facet_wrap(~year) +
  geom_errorbar(aes(x = prop_responses_mean, y = sample_size_mean, color = response, ymin = sample_size_mean - sample_size_sd, ymax = sample_size_mean + sample_size_sd)) +
  geom_errorbarh(aes(x = prop_responses_mean, y = sample_size_mean, color = response, xmin = prop_responses_mean - prop_responses_sd, xmax = prop_responses_mean + prop_responses_sd)) +
  labs(title = "Average Sample Size and Proportion of Responses in Minnesota by Year", x = "Average Proportion of Responses", y = "Average of Sample Size", color = "Response")
```
This plot shows the average and standard deviation of the sample size and the average and standard deviation of the proportions for each response in Minnesota for the years 2002, 2006, and 2010. 

We can also use the figures below if we want to get a better look at the means and standard deviations for proportion and sample size.


```{r, message=F, warning=F}
tidy_brfss %>%
  filter(year %in% c(2002, 2006, 2010), state %in% "MN") %>%
  group_by(year, response) %>%
  summarize(across(c(sample_size, prop_responses), list(mean = mean,sd = sd))) %>%
  ggplot(aes(x = response, y = prop_responses_mean, fill = response)) +
  geom_col() +
  facet_wrap(~year) +
  geom_errorbar(aes(x = response, y = prop_responses_mean, fill = response, ymin = prop_responses_mean - prop_responses_sd, ymax = prop_responses_mean + prop_responses_sd), position = "dodge") +
  theme(axis.text.x = element_text(angle = 90)) +
  labs(title = "Average Proportion of Respones in Minnesota by Year", x = "Response", y = "Average of Proportion", fill = "Response")
```

```{r, message=F, warning=F}
tidy_brfss %>%
  filter(year %in% c(2002, 2006, 2010), state %in% "MN") %>%
  group_by(year, response) %>%
  summarize(across(c(sample_size, prop_responses), list(mean = mean,sd = sd))) %>%
  ggplot(aes(x = response, y = sample_size_mean, fill = response)) +
  geom_col() +
  facet_wrap(~year) +
  geom_errorbar(aes(x = response, y = sample_size_mean, fill = response, ymin = sample_size_mean - sample_size_sd, ymax = sample_size_mean + sample_size_sd), position = "dodge") +
  theme(axis.text.x = element_text(angle = 90)) +
  labs(title = "Average Sample Size of Responses in Minnesota by Year", x = "Response", y = "Average of Sample Size", fill = "Response")
```